Redshift evolution of the SN stretch distribution

### 2019-10-21 Nicolas.tex
Results. Need to add table pages. Preambule shortened.

The simple use of those is not enough for our aim, as SNe Ia have an intrinsec
dispersion of their luminosity of $\approx \SI{0.4}{mag}$ that gives a huge
uncertainty on the determination of their distance modulus. Henceforth they are
standardized using the "brigther-slower" and "brighter-bluer" relations
(Philipps 93, Riess 96, Tripp 98) in the SALT2 algorithm (Guy 2007, 2010) that
fits the distance modulus which is expressed as
\begin{center}
    $ \displaystyle \m = m_b + \a x_1 - \b c - M$
\end{center}
with $m_B$ the logarithm of their flux, $x_1$ their stretch, $c$ their color and
$M$ their intrinsec magnitude. This relation lowered their magnitude dispersion
to $\approx \SI{0.15}{mag}$, allowing for previously mentioned accelerated
expansion to be discovered.

@@

As such, we determined the maximum redshift at which we would cut these samples
to get complete ones. We used a statistical approach in lack of precise data
concerning the instruments' capacity to acquire fluxes. We proceeded by making
histograms of the surveys, and defining a function of the expected SNe Ia rate:
\begin{center}
    $ \displaystyle N\St{SNe Ia} = a \times V\z$
\end{center}
with $a$ the density of SNe Ia and $V\z$ the volume of Universe in each bin.
Before selection effects, the histograms are supposed to grow proportionnaly to
$V\z$ assuming an homogeneous repartition of SNe Ia. When we start to miss some
supernovae, this growth is not followed anymore.

To determine this limit, we first took the observed counts in each bin of the
histograms. We then fitted $a$ on these counts for different numbers of bins
used to fit: the quality of the fit diminishes when we use bins that include
selection effects. To quantize this, we computed the expected counts from the
fitted $a$ in the same bins we used for the observed counts. That allowed us to
compute the poissonian probability in each bin, given by
\begin{center}
    $ \displaystyle p(k) = \frac{\l^k}{k!}e^{-\l}$
\end{center}
with $k$ the expected counts, $\l$ the observed counts. This work is done 100
times with a random number of total bins, $a$ being fitted 10 times on a
different and random number of bins used to fit each time. We interpolate these
1000 poissonian evolutions to get one final result that is as close as possible
to accurately represent how the probability to follow the expected rate evolves.

@@

We used the LsSFR and stretch measurments from the SNf sample. The LsSFR has an
evolution with the redshift that is analytically known: calling $\de\z$ the
fraction of young stars and $\p\z$ the fraction of old ones, Rigault 2018 et +
find
\begin{equation}
    \frac{\de\z}{\p\z} \equiv LsSFR\z = K\times \LF{1+z}^\f
\end{equation}
with $\f = 2.8$, and knowing $\de\z + \p\z = 1$:
\begin{align}
    \de\z & = \LF{K^{-1}\times \LF{1+z}^{-\f} +1}^{-1}\\
    \p\z & = \LF{K\times \LF{1+z}^{+\f} + 1}\um
\end{align}

The goal is to find how the stretch depends on the LsSFR, distinguishing old and
young SNe which fractions evolve with the redshift, in order to have an
analytical law for the mean redshift evolution of the stretch. The measured data
on which we based our first model is plotted in figure \ref{fig:mod_first};
based of the shape of the $x_1$ vs LsSFR cloud, we implemented the following
model:
\begin{itemize}
    \item \textbf{young}: a gaussian of mean $\m_1$ and standard deviation
        $\s_1$, namely $\Nc_1 \equiv \Nc \LF{\m_1,\s_1}$;
    \item \textbf{old}: a linear combination between $\Nc_1$ and another
        gaussian $\Nc_2 \equiv \Nc \LF{\m_2,\s_2}$
\end{itemize}
As such, the probability to observe a young SNe Ia labeled "i" with a stretch
$x_1^i$ and error $\d x_1^i$ is
\begin{equation}
    p \LF{x_1^i, \d x_1{}^i|\m_1,\s_1} = \Nc \LF{\m_1, \Sq{\s_1{}^2 + \d
    x_1^i{}^2}}(x_1^i)
\end{equation}
and for an old one:
\begin{align}
    p(x_1^i, \d x_1^i | \m_{1,2}, \s_{1,2}, a) = a & \times \Nc \LF{\m_1,
    \sqrt{\s_1{}^2 + \d x_1^i{}^{2}}} (x_1^i) \ + \\ (1-a) & \times \Nc
    \LF{\m_2, \sqrt{ \s_2{}^2 + \d x_1^i{}^{2}}} (x_1^i), \nonumber 
\end{align}
We fit different model, and compare their AICc.

@@

\section{Results}
We find that every model lacking an evolution of the stretch with the redshift
is systematically worse than those that implement it.

>>>>

The simple use of those is not enough for our aim, as SNe Ia have an intrinsec
dispersion of their luminosity of $\approx 0.4~mag$ that gives a huge
uncertainty on the determination of their distance modulus. Henceforth they are
standardized using the "brigther-slower" and "brighter-bluer" relations
(Philipps 93, Riess 96, Tripp 98) in the SALT2 algorithm (Guy 2007, 2010) that
fits the distance modulus which is expressed as
\begin{equation}
    \m = m_b + \a x_1 - \b c - M
\end{equation}
with $m_B$ the logarithm of their flux, $x_1$ their stretch, $c$ their color and
$M$ their intrinsec magnitude. This relation lowered their magnitude dispersion
to $\approx 0.15~mag$, allowing for previously mentioned accelerated
expansion to be discovered.

@@

As such, we determined the maximum redshift at which we would cut these samples
to get complete ones. We used a statistical approach in lack of precise data
concerning the instruments' capacity to acquire fluxes. We proceeded by making
histograms of the surveys, and defining a function of the expected SNe Ia rate:
\begin{equation}
    N\St{SNe Ia} = a \times V\z
\end{equation}
with $a$ the density of SNe Ia and $V\z$ the volume of Universe in each bin.
Before selection effects, the histograms are supposed to grow proportionnaly to
$V\z$ assuming an homogeneous repartition of SNe Ia. When we start to miss some
supernovae, this growth is not followed anymore.

To determine this limit, we first took the observed counts in each bin of the
histograms. We then fitted $a$ on these counts for different numbers of bins
used to fit: the quality of the fit diminishes when we use bins that include
selection effects. To quantize this, we computed the expected counts from the
fitted $a$ in the same bins we used for the observed counts. That allowed us to
compute the poissonian probability in each bin, given by
\begin{equation}
    p(k) = \frac{\l^k}{k!}e^{-\l}
\end{equation}
with $k$ the expected counts, $\l$ the observed counts. This work is done 100
times with a random number of total bins, $a$ being fitted 10 times on a
different and random number of bins used to fit each time. We interpolate these
1000 poissonian evolutions to get one final result that is as close as possible
to accurately represent how the probability to follow the expected rate evolves.

@@

We used the LsSFR and stretch measurments from the SNf sample. The LsSFR has an
evolution with the redshift that is analytically known: calling $\de\z$ the
fraction of young stars and $\p\z$ the fraction of old ones, Rigault 2018 et +
find
\begin{equation}
    \frac{\de\z}{\p\z} \equiv \mathrm{LsSFR}\z = K\times \LF{1+z}^\f
\end{equation}
with $\f = 2.8$, and knowing $\de\z + \p\z = 1$:
\begin{align}
    \de\z & = \LF{K^{-1}\times \LF{1+z}^{-\f} +1}^{-1}\\
    \p\z & = \LF{K\times \LF{1+z}^{+\f} + 1}\um
\end{align}

@@

The goal is to find how the stretch depends on the LsSFR, distinguishing old and
young SNe which fractions evolve with the redshift, in order to have an
analytical law for the mean redshift evolution of the stretch. The measured data
on which we based our first model is plotted in figure \ref{fig:mod_first};
based of the shape of the $x_1$ vs LsSFR cloud, we implemented the following
model:
\begin{itemize}
    \item \textbf{young}: a gaussian of mean $\m_1$ and standard deviation
        $\s_1$, namely $\Nc_1 \equiv \Nc \LF{\m_1,\s_1}$;
    \item \textbf{old}: a linear combination between $\Nc_1$ and another
        gaussian $\Nc_2 \equiv \Nc \LF{\m_2,\s_2}$
\end{itemize}
As such, the probability to observe a young SNe Ia labeled "i" with a stretch
$x_1^i$ and error $\d x_1^i$ is
\begin{equation}
    p \LF{x_1^i, \d x_1{}^i|\m_1,\s_1} = \Nc \LF{\m_1, \Sq{\s_1{}^2 + \d
    x_1^i{}^2}}(x_1^i)
\end{equation}
and for an old one:
\begin{align}
    p(x_1^i, \d x_1^i | \m_{1,2}, \s_{1,2}, a) = a & \times \Nc \LF{\m_1,
    \sqrt{\s_1{}^2 + \d x_1^i{}^{2}}} (x_1^i) \ + \\ (1-a) & \times \Nc
    \LF{\m_2, \sqrt{ \s_2{}^2 + \d x_1^i{}^{2}}} (x_1^i), \nonumber 
\end{align}

\noindent where $a$ is the relative amplitude between the two gaussians.
Finally, the normalized stretch distribution at a given $z$ $\D\LF{x_1|z}$ is
the weigthed sum of both young and old stretch distributions given their
relative fraction $\de\z$:
\begin{equation}
    \D\LF{x_1|z} = \de\z\times\Nc_1 +
    \LF{1-\de\z} \times \LF{a\Nc_1 + (1-a)\Nc_2}
\end{equation}

We fitted it on SNf data, giving the results table \ref{tab:val}. For clarity
with the next models, we named it 3G2M2S$\St{SNf}$ for it has a total of 3
gaussians but with only 2 means and 2 standard deviations, and has been fitted
on SNf data only.

We implemented and compared 10 models in total, 4 of which have an
evolution with the redshift from $\de\z$, and 6 don't ($\de\z = f =
\mathrm{cst}$). The ones with and evolution are:
\begin{itemize}
    \item 3G2M2S, the one we described (but fitted on all the data);
    \item 3G2M1S, where this time $\s_1 \equiv \s_2$;
    \item 2G2M2S, model taken from HOWELL 2009 where we added $\de\z$;
    \item 3G3M3S, with three independent gaussians.
\end{itemize}
The ones without a stretch evolution are the same ones but with an "F" implying
we set $\de\z = f = \mathrm{cst}$, and two others:
\begin{itemize}
    \item 1G1M1S, where there is no distinction between old and young SNe;
    \item 1G1M2S, taken from KESLLER 2017 and used in recent cosmological
        analysis SCOLNIC 2018. It's an asymetric model where
\end{itemize}

\begin{align}
 p(x_1^i, \d x_1^i | \m, \s_-, \s_+) = 
    \begin{cases}
        \mathcal{N} \LF{\m, \sqrt{\s_-{}^2 + \d x_1^i{}^{2}}} (x_1^i) & \text{if
        } x_1^i\geq \m\\
        \mathcal{N} \LF{\m, \sqrt{\s_+{}^2 + \d x_1^i{}^{2}}} (x_1^i), &
        \text{else}
    \end{cases}
\end{align}

The fitted parameters are showed table \ref{tab:val}.

\begin{table}[htbp!]
    \centering
    \caption{Valeurs des paramètres pour différents modèles. En rouge les
    données aberrantes.}
    \label{tab:val}
    \begin{tabularx}{\linewidth}{|Y*{6}{|Y}|}\hline

        \rowcolor{gray!15} Modèle & $a$ & $f$ & $\m_1$ & $\s_1$ & $\m_2$ &
        $\s_2$ \\\hline\hline

        3G2M2S$_{\mathrm{SNf}}$ & $0.48 \pm 0.07$ & none & $0.39 \pm 0.07$ &
        $0.56 \pm 0.05$ & $-1.5 \pm 0.1$ & $0.52 \pm 0.09$ \\\hline

        3G2M2S & $0.48 \pm 0.17$ & none & $0.36 \pm 0.08$ & $0.61 \pm 0.05$ &
        $-1.3 \pm 0.2$ & $0.60 \pm 0.12$ \\\hline

        3G2M2SF & \textcolor{red}{$0.1 \pm 0.6$} & \textcolor{red}{$0.2 \pm
        0.6$} & $-0.9 \pm 0.7$ & $0.7 \pm 0.3$ & $0.5 \pm 0.2$ & $0.6 \pm 0.1$
        \\\hline

        3G2M1S & $0.47 \pm 0.07$ & none & $0.35 \pm 0.04$ & $0.61 \pm 0.03$ &
        $-1.25 \pm 0.10$ & $\s_1$ \\\hline

        3G2M1SF & \textcolor{red}{$0.2 \pm 0.9$} & $0.7 \pm 0.3$ & $0.36 \pm
        0.04$ & $0.60 \pm 0.03$ & $-1.23 \pm 0.10$ & $\s_1$ \\\hline

        2G2M2S & none & none & $0.49 \pm 0.04$ & $0.54 \pm 0.03$ & $-0.72 \pm
        0.08$ & $0.83 \pm 0.07$ \\\hline

        2G2M2SF & none & $0.3 \pm 0.2$ & $-0.9 \pm 0.6$ & $0.7 \pm 0.2$ & $0.5
        \pm 0.2$ & $0.56 \pm 0.09$ \\\hline

    \end{tabularx} \bigbreak

\begin{tabularx}{\linewidth}{|Y*{2}{|Y}|}\hline

    \rowcolor{gray!15} Modèle & $\m$ & $\s$ \\\hline\hline

    1G1M1S & $0.01 \pm 0.04$ & $0.90 \pm 0.03$ \\\hline

\end{tabularx} \bigbreak

\begin{tabularx}{\linewidth}{|Y*{3}{|Y}|}\hline

    \rowcolor{gray!15} Modèle & $\m$ & $\s_-$ & $\s_+$ \\\hline\hline

    1G1M2S & $0.16617 \pm 0.00004$ & $1.07 \pm 0.04$ & $0.69 \pm 0.03$ \\\hline

\end{tabularx} \bigbreak

\begin{tabularx}{\linewidth}{|Y*{8}{|Y}|}\hline

    \rowcolor{gray!15} Modèle & $a$ & $f$ & $\m_1$ & $\s_1$ & $\m_2$ & $\s_2$ &
    $m_3$ & $\s_3$ \\\hline\hline

    3G3M3S & $0.14 \pm 0.08$ & none & $0.51 \pm 0.06$ & $0.54 \pm 0.04$ & $-1.9
    \pm 0.2$ & $0.29 \pm 0.11$ & $-0.55 \pm 0.12$ & $0.67 \pm 0.15$ \\\hline

    3G3M3SF & $0.2 \pm 0.2 $ & $0.10 \pm 0.04 $ & $-1.7 \pm 0.2$ & $0.4 \pm 0.1$
            & $0.9 \pm 0.1$ & $0.3 \pm 0.2$ & $0.0 \pm 0.2$ & $0.7 \pm 0.1$
            \\\hline

\end{tabularx} \bigbreak

\end{table}

\section{Results}
To compare all these models, we use the Akaike Information Criterion corrected
for sample size (BURNHAM 2002), that penalises the increase of free parameters
in order to discourage overfitting:
\begin{equation}
    \mathrm{AICc} = \mathrm{AIC} + \LR{2k\LF{k+1}}{n - k - 1}
\end{equation}
with $\mathrm{AIC} = 2k - 2\ln\LF{\mathcal{L}}$, where $k$ is the number of free
parameters and $\Lc$ the likelihood. The probability for a model to be as
representative as the "best" one is given by:
\begin{equation}
    p(\mathrm{other} > \mathrm{best}) = \exp\LF{\D\mathrm{AICc}/2}
\end{equation}
As such, we obtain the results table \ref{tab:comp}.

\begin{table}[htbp!]
    \centering
    \caption{Comparaison des modèles. NR représente les modèles implémentés
             durant ce stage. (F) indique les modèles pour lesquels il n'y a pas
             d'évolution de la fraction de SNe~Ia jeunes et vieilles en fonction
             du redshift.}
    \label{tab:comp}
    \begin{tabularx}{\linewidth}{|Y*{6}{|Y}|}\hline

        \rowcolor{gray!15} Name & Description & Free param &
        $\ln\mathcal{L}$ & $AICc$ & $\D AICc$ & Proba \\\hline\hline

        3G2M1S & NR 1S & 4 & 1815 & 1823 & 0.0 & $1.0$ \\\hline

        3G2M2S & NR 2S & 5 & 1815  & 1825 & -2.0 & $3.6\times10^{-1}$ \\\hline

        2G2M2S & Howell & 4 & 1818  & 1826 & -3.4 & $1.8\times10^{-1}$ \\\hline

        3G3M3S & NR 3S & 7 & 1812  & 1826 & -3.6 & $1.6\times10^{-1}$ \\\hline

        3G3M3SF & NR 3S (F) & 8 & 1813 & 1829 & -6.3  & $4.3\times10^{-2}$
        \\\hline

        2G2M2SF & Howell (F) & 5 & 1823  & 1833 & -9.9 & $7.0\times10^{-3}$
        \\\hline

        3G2M1SF & NR 1S (F) & 5 & 1823  & 1833 & -10.4 & $5.5\times10^{-3}$
        \\\hline

        3G2M2SF & NR 2S (F) & 6 & 1823  & 1835 & -12.0 & $2.5\times10^{-3}$
        \\\hline

        1G1M2S & Kessler (F) & 3 & 1837  & 1843 & -20.2 & $4.1\times10^{-5}$
        \\\hline

        1G1M1S & 1 gauss. (F) & 2 & 1872 & 1876 & -53.5 &
        $2.4\times10^{-12}$\\\hline

    \end{tabularx}
\end{table}

We find that every model lacking an evolution of the stretch with the redshift
is systematically worse than those that implement it.
